---
title: "Multivariate O-statistics"
author: "Quentin D. Read, (INSERT OTHER AUTHORS OF THIS PARTICULAR DOCUMENT)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multivariate-Ostats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Ostats)
data(pitcher_traits)
```

# Notes

- Remove this section in the final version.
- Authors of this doc aren't necessarily all package authors, just those who specifically contributed to this vignette.
- Outline of the vignette:
  + Explain multivariate O-stats, using hypervolumes
  + Show a "toy" version of the pitcher plant dataset
  + Run a few different variants of multivariate O-stats on them
  + Create some plots
  
# Overview

This vignette is intended for users who are already familiar with the single-variable O-statistics we introduce in the first package vignette (LINK TO IT).

- Hutchinson: niche is n-dimensional hypervolume
- trait ecologists are interested in tradeoffs among traits or interactions among multiple traits
- the underlying math and theory here were worked out by Blonder et al.
- Functions from their hypervolume package are used in Ostats

# Example dataset

- Pitcher plant traits, morphological measurements taken from sites around North America
- Specific site locations have been removed
- Univariate plots of pitcher traits (distributions)

# Calculating hypervolume O-statistics

- Calculate ostats multivariate 
- Do a couple different versions with different arguments passed to hypervolume() to demonstrate the sensitivity of results to those args

# Visualizing hypervolume O-statistics

- Show the ostats multivariate plot function
- Show a couple different plots to illustrate the effects of different arguments

# Some practical notes

- Warning that it's important to make sure there aren't a lot of correlated traits in your dataset or you will end up with spurious hypervolumes
- Absolute amount of overlap can't be compared among hypervolumes with different dimensionality, so it's especially important to run null models
