---
title: "Multivariate O-statistics"
author: "Quentin D. Read, (INSERT OTHER AUTHORS)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multivariate-Ostats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Ostats)
data(pitcher_traits)
```

# Notes

- Remove this section in the final version.
- Authors of this doc aren't necessarily all package authors, just those who specifically contributed to this vignette.
- Outline of the vignette:
  + Explain multivariate O-stats, using hypervolumes
  + Show a "toy" version of the pitcher plant dataset
  + Run a few different variants of multivariate O-stats on them
  + Create some plots
  
# Overview

This vignette is intended for users who are already familiar with the single-variable O-statistics we introduce in the first package vignette (LINK TO IT).

- Hutchinson: niche is n-dimensional hypervolume
- trait ecologists are interested in tradeoffs among traits or interactions among multiple traits
- the underlying math and theory here were worked out by Blonder et al.
- Functions from their hypervolume package are used in Ostats

# Example dataset

- Pitcher plant traits, morphological measurements taken from sites around North America
- Specific site locations have been removed
- Univariate plots of pitcher traits (distributions)

Here we show a few randomly selected rows from the pitcher traits dataset. Here, all individuals are of the same species. Therefore we will use populations (sites) as the `sp` variable in our analysis, with a single value for `plots`. 

```{r}
set.seed(1)
idx <- sort(sample(nrow(pitcher_traits), 10, replace = FALSE))

knitr::kable(pitcher_traits[idx,])
```

# Calculating hypervolume O-statistics

- Calculate ostats multivariate 
- Do a couple different versions with different arguments passed to hypervolume() to demonstrate the sensitivity of results to those args

Here we calculate the univariate O-statistics for each of the pitcher traits separately, then a single multivariate O-statistic representing overlap of the trait hypervolumes. 

We will use four morphological ratios as traits to remove size-dependence: the ratios of pitcher length, keel width, mouth diameter, and lip thickness to pitcher width.

First ensure that the data are in the appropriate form for `Ostats()` and `Ostats_multivariate()`. For this example we will remove individuals with incomplete data.

```{r}
pitcher_traits <- pitcher_traits[complete.cases(pitcher_traits), ]

pitcher_sites <- as.factor(pitcher_traits$site_id)

pitcher_trait_matrix <- as.matrix(pitcher_traits[, c('pitcher_length', 'keel_width', 'mouth_diameter', 'lip_thickness')])

pitcher_trait_matrix <- sweep(pitcher_trait_matrix, 1, pitcher_traits$pitcher_width, `/`)

```

The traits are all on a ratio scale but we should center and scale all the columns to ensure that the hypervolume overlap statistic is not influenced by different magnitudes of each trait.

```{r}
apply(pitcher_trait_matrix, 2, range)

pitcher_trait_matrix_scaled <- scale(pitcher_trait_matrix, center = TRUE, scale = TRUE)

apply(pitcher_trait_matrix_scaled, 2, range)
```


Next calculate the univariate O-statistics for each of the four traits. **Currently in the example we will not use null model, but eventually it should.**

```{r, message = FALSE}
pitcher_univariate <- Ostats(traits = pitcher_trait_matrix_scaled,
                             plots = factor(rep(1, nrow(pitcher_trait_matrix_scaled))),
                             sp = pitcher_sites,
                             random_seed = 444,
                             run_null_model = FALSE
)
```

Now calculate the multivariate O-statistic for the traits combined, using all default arguments. The `random_seed` is still used even in the absence of a null model for the hypervolume construction.

```{r, message = FALSE}
pitcher_multivariate <- Ostats_multivariate(traits = pitcher_trait_matrix_scaled[, 1:3],
                                            plots = factor(rep(1, nrow(pitcher_trait_matrix_scaled))),
                                            sp = pitcher_sites,
                                            random_seed = 333,
                                            run_null_model = FALSE
)
```
Compare the univariate and multivariate O-statistics.

```{r}
pitcher_univariate$overlaps_norm

pitcher_multivariate$overlaps_norm
```


**Here run the multivariate O-stats with different args to hypervolume**

# Visualizing hypervolume O-statistics

- Show the ostats multivariate plot function
- Show a couple different plots to illustrate the effects of different arguments

Here is a plot of the overlap of the hypervolumes in 4-dimensional space formed by the individual traits of the pitcher plants at each site. Notice that we don't use the scaled traits for plotting; instead we use the actual ratios so that the axes are interpretable. 

```{r, message = FALSE}
Ostats_multivariate_plot(plots = factor(rep(1, nrow(pitcher_trait_matrix))),
                         sp = pitcher_sites,
                         traits = pitcher_trait_matrix,
                         overlap_dat = pitcher_multivariate,
                         plot_points = TRUE 
)
```


# Some practical notes

- Warning that it's important to make sure there aren't a lot of correlated traits in your dataset or you will end up with spurious hypervolumes
- Absolute amount of overlap can't be compared among hypervolumes with different dimensionality, so it's especially important to run null models
- Biologically, it's been shown that the multi-trait stuff does not contribute that much to what we can infer. This may not be worth discussing in the vignette, but should be discussed in the paper.
