---
title: "Multivariate O-statistics"
author: "Quentin D. Read, (INSERT OTHER AUTHORS)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multivariate-Ostats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png", 
  dev.args = list(type = "cairo-png")
)
```

```{r setup}
library(Ostats)
data(pitcher_traits)
```

# Notes

- Remove this section in the final version.
- Authors of this doc aren't necessarily all package authors, just those who specifically contributed to this vignette.
- Outline of the vignette:
  + Explain multivariate O-stats, using hypervolumes
  + Show a "toy" version of the pitcher plant dataset
  + Run a few different variants of multivariate O-stats on them
  + Create some plots
  
# Overview

This vignette is intended for users who are already familiar with the single-variable O-statistics we introduce in the first package vignette (LINK TO IT).

- Hutchinson: niche is n-dimensional hypervolume
- trait ecologists are interested in tradeoffs among traits or interactions among multiple traits
- the underlying math and theory here were worked out by Blonder et al.
- Functions from their hypervolume package are used in Ostats

# Example dataset

- Pitcher plant traits, morphological measurements taken from sites around North America
- Specific site locations have been removed
- Univariate plots of pitcher traits (distributions)

Here we show a few randomly selected rows from the pitcher traits dataset. Here, all individuals are of the same species. Therefore we will use populations (sites) as the `sp` variable in our analysis, with a single value for `plots`. 

```{r}
set.seed(1)
idx <- sort(sample(nrow(pitcher_traits), 10, replace = FALSE))

knitr::kable(pitcher_traits[idx,])
```

For a more manageable example, let's select 6 sites.

```{r}
sites <- c('COL', 'CRI', 'FLK', 'FRA', 'MYR', 'SAC')
pitcher_traits <- pitcher_traits[pitcher_traits$site_id %in% sites, ]
```


# Calculating hypervolume O-statistics

- Calculate ostats multivariate 
- Do a couple different versions with different arguments passed to hypervolume() to demonstrate the sensitivity of results to those args

Here we calculate the univariate O-statistics for each of the pitcher traits separately, then a single multivariate O-statistic representing overlap of the trait hypervolumes. 

When estimating hypervolumes, as with any other multivariate analysis, it is critical to ensure that the hypervolume axes are not too highly correlated with one another. If traits are too highly correlated, the calculated hypervolume will not accurately represent the distribution in trait space. In this case, we resolve the issue by excluding a few of the traits with high correlations. You could also choose to use a variance maximization procedure like PCA instead.

First ensure that the data are in the appropriate form for `Ostats()` and `Ostats_multivariate()`. For this example we will remove individuals with incomplete data. Using `cor()` on the cleaned dataset, we can see the relatively high correlations among some of the traits.

```{r}
pitcher_traits <- pitcher_traits[complete.cases(pitcher_traits), ]

pitcher_sites <- as.factor(pitcher_traits$site_id)

round(cor(pitcher_traits[, -(1:2)]), 2)
```

We select four traits that represent dimensions of different parts of the plant's anatomy. As a result they are relatively uncorrelated with one another: rosette diameter, pitcher width, mouth diameter, and lip thickness.

```{r}
traits_to_use <- c("rosette_diameter_1", "pitcher_width", "mouth_diameter", "lip_thickness")

pitcher_trait_matrix <- as.matrix(pitcher_traits[, traits_to_use])

round(cor(pitcher_trait_matrix), 2)
```

Next we center and scale all the columns to ensure that the hypervolume overlap statistic is not influenced by different magnitudes of each trait.

```{r}
apply(pitcher_trait_matrix, 2, range)

pitcher_trait_matrix_scaled <- scale(pitcher_trait_matrix, center = TRUE, scale = TRUE)

round(apply(pitcher_trait_matrix_scaled, 2, range), 2)
```


Next calculate the univariate O-statistics for each of the four traits. 

```{r univariate O-stats, message = FALSE, results = 'hide'}
pitcher_univariate <- Ostats(traits = pitcher_trait_matrix_scaled,
                             plots = factor(rep(1, nrow(pitcher_trait_matrix_scaled))),
                             sp = pitcher_sites,
                             random_seed = 444,
                             run_null_model = FALSE
)
```

Now calculate the multivariate O-statistic for the traits combined, using all default arguments. The `random_seed` is still used even in the absence of a null model because the hypervolumes are constructed with a stochastic algorithm.

```{r multivariate O-stats, message = FALSE, results = 'hide'}
pitcher_multivariate <- Ostats_multivariate(traits = pitcher_trait_matrix_scaled,
                                            plots = factor(rep(1, nrow(pitcher_trait_matrix_scaled))),
                                            sp = pitcher_sites,
                                            random_seed = 333,
                                            run_null_model = FALSE
)
```
Compare the univariate and multivariate O-statistics.

```{r}
pitcher_univariate$overlaps_norm

pitcher_multivariate$overlaps_norm
```

The single multivariate O-statistic is not comparable to the univariate statistics because of differing dimensionality. Therefore we need to evaluate them against null models.

**Here run the multivariate O-stats with different args to hypervolume, and with null models**

# Visualizing hypervolume O-statistics

- Show the ostats multivariate plot function
- Show a couple different plots to illustrate the effects of different arguments
- Give credit to hypervolume package for the plot.Hypervolumelist function

We can visualize the overlap of the different hypervolumes using the function `Ostats_multivariate_plot()`. This function returns a list of plot objects, one for each community. Each plot displays an upper triangular matrix of panels, one for each pair of traits, with the cross-sections of the hypervolumes in those two dimensions, as well as the raw data points.

For example, we can plot the overlap of the hypervolumes in 4-dimensional space formed by the individual traits of the pitcher plants at each site. ~~Notice that we don't use the scaled traits for plotting; instead we use the raw trait values so that the axes are interpretable.~~ 

```{r default plot, message = FALSE, fig.height = 9, fig.width = 9}
Ostats_multivariate_plot(plots = factor(rep(1, nrow(pitcher_trait_matrix))),
                         sp = pitcher_sites,
                         traits = pitcher_trait_matrix_scaled,
                         overlap_dat = pitcher_multivariate,
                         plot_points = TRUE
)
```

We can customize the plot by supplying a color palette to the `colorvalues` argument (the default is a colorblind-friendly palette imported from [viridis](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)), and add additional spacing around the hypervolume contours using the `axis_buffer_factor` argument. The default value of this argument is `0.25`, or 25% of the range of the data in all directions. In some cases adding additional spacing may be necessary because the contours may extend farther than that. Finally, we can suppress plotting the raw data and only plot the contours using `plot_points = FALSE`.

FIXME: in the actual function we should include argument to buffer the plot itself, not just the contour estimation.

```{r custom plot, message = FALSE, fig.height = 9, fig.width = 9}
Ostats_multivariate_plot(plots = factor(rep(1, nrow(pitcher_trait_matrix))),
                         sp = pitcher_sites,
                         traits = pitcher_trait_matrix,
                         overlap_dat = pitcher_multivariate,
                         colorvalues = rainbow(length(sites)),
                         axis_buffer_factor = 0.5,
                         plot_points = FALSE
)
```

# Some practical notes

- Warning that it's important to make sure there aren't a lot of correlated traits in your dataset or you will end up with spurious hypervolumes
- Absolute amount of overlap can't be compared among hypervolumes with different dimensionality, so it's especially important to run null models
- Biologically, it's been shown that the multi-trait stuff does not contribute that much to what we can infer. This may not be worth discussing in the vignette, but should be discussed in the paper.
